Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond

### 2024-04-09 main.tex
:pencil: Update contributions

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables {cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} in the Stiefel manifold, which performs multiplicative (instead of add-then-retract) updates along the exponential map for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	For the singular value problem, we characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of {subchannel rearrangement} and {subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

\begin{section}{Introduction}
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable metasurface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than relays, and greater scalability than \gls{mimo}.
	The most popular \gls{ris} application is \emph{joint beamforming} design with transceivers, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	Since coherent scattering can be decomposed as equal-gain combining and transmission, \gls{ris} elements usually offer a squared asymptotic performance (e.g., second-order array gain and fourth-order power scaling law \cite{Zhao2022}) than transceiver antennas.
	% However, a large-dimensional \gls{ris} is generally required to compensate the power loss from double fading.
	On the other hand, \gls{ris} can also be employed for \emph{backscatter modulation} by periodically switching its reflection pattern.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) that can be either integrated into the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or exploited by surrounding low-power devices for energy-efficient uplink communications \cite{Liang2020,Zhao2022a,Yang2024}.

	Despite the great potentials and fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape a \gls{mimo} channel?}
	The answer depends heavily on the scattering model and hardware architecture.
	In diagonal \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	This ideally translates to a diagonal scattering matrix with unit-magnitude entries, which only controls the phase of the scattered waves but cannot redistribute their amplitude.
	The concept was soon generalized to \gls{bd}-\gls{ris} \cite{Shen2020a} which physically groups adjacent elements using passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices respectively.}
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate losslessly within the circuit and depart partially from any element in the same group.
	It can thus control both the amplitude and phase of the scattered wave without power loss, generalizing the scattering matrix to block-diagonal with unitary blocks.
	Such a powerful model can be realized at reduced hardware complexity using tree- and forest-connected architectures inspired by graph theory \cite{Nerini2024}.
	\gls{bd}-\gls{ris} can also function in hybrid transmitting-and-reflecting mode \cite{Li2023b} and multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
	Many practical design challenges have been addressed including channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a}.
	Its superiority has been studied extensively in \gls{siso} and \gls{miso} systems in terms of single-user \gls{snr} maximization \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and multi-user \gls{wsr} maximization \cite{Fang2023,Zhou2023,Li2023c,Soleymani2024}.
	When it comes to \gls{mimo}, most existing works have been limited to very specific scenarios.
	For example, the authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design under the assumption of unitary \gls{ris} without direct link.
	Received power maximization problem is extended to \gls{mimo} for continuous-valued \cite{Nerini2023} and discrete-valued \cite{Nerini2023b} scattering matrices, but they only considered single-stream transmission with given precoder and combiner (which is equivalent to \gls{siso}).
	Transmitter-side \gls{bd}-\gls{ris} design over statistical \gls{csi} has been introduced to massive \gls{mimo} networks for improved spectral efficiency \cite{Mishra2024}, but was again constrained to unitary \gls{ris} without direct link.
	A practical frequency-dependent \gls{bd}-\gls{ris} model has been recently proposed for multi-band multi-cell \gls{mimo} networks to facilitate practical deployments \cite{Sena2024}.
	Although the results are insightful, it remains unclear how the off-diagonal entries contribute to channel shaping and whether the relative gain of \gls{bd}-\gls{ris} scales with \gls{mimo} dimensions.
	% Therefore, \gls{bd}-\gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	Channel shaping is different from passive beamforming since the \gls{ris} is exploited as a stand-alone device to modify the inherent properties of the channel itself, instead of optimizing for a specific performance metric.
	This not only allows one to decouple the joint \gls{ris}-transceiver design, but also unveils the fundamental limits of wave manipulation and provides an all-in-one benchmark for various wireless applications.
	Relevant designs can be classified into two categories:
	\begin{itemize}
		\item \emph{Singular value centric:} Focus on the singular values of the equivalent channel matrix, which are sensitive to minor perturbations but precisely related to the capacity and diversity-multiplexing trade-off. It has been mainly studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}.
		\item \emph{Power centric:} Focus on the second-order characteristics of the equivalent channel matrix, which are useful in \gls{siso} and \gls{miso} but less informative in \gls{mimo}. It has been mainly studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and leakage interference \cite{Santamaria2023a}.
	\end{itemize}

	While the above works offer promising glimpses into the channel shaping potential of passive \gls{ris}, the big picture remains incomplete because they neither suggest the achievable singular value region nor address the most general \gls{bd}-\gls{ris}-aided \gls{mimo} scenario.
	On the other hand, the great potential of \gls{bd}-\gls{ris} has not been fully understood and an efficient design framework is still missing.
	This chapter aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	First, we exploit the Riemannian geometry of the Stiefel manifold to design a novel geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg} algorithm for \gls{bd}-\gls{ris} optimization.
	This method modified from \cite{Abrudan2008,Abrudan2009} not only features lower complexity and faster convergence than general non-geodesic approach \cite{Absil2009,Pan2022d}, but also fits arbitrary group size and any smooth optimization problem.
	Specifically, block-wise multiplicative rotational updates are performed along the geodesics of the Stiefel manifold, which are compactly evaluated as the exponential map \cite{Edelman1998}.
	By exploiting the inherent structure of unitary matrices, this method avoids retractions from the Euclidean space and improves the computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

	Second, we demystify the gain from off-diagonal entries of \gls{bd}-\gls{ris} in terms of subchannel rearrangement and subspace alignment.

	% Specifically,  we show that BD-RIS enables cooperative groups to rearrange subchannels according to their strengths, maximizing the effective channel gain. Additionally, intra-group cooperation allows for finer control over subspace alignment compared to diagonal RIS, especially in higher-dimensional MIMO systems. This analysis highlights a key advantage of BD-RIS and its potential for significant performance improvement.

	Third, we quantify the capability of a \gls{bd}-\gls{ris} to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The \emph{Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The resulting achievable singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and fully-connected (a.k.a. unitary) \gls{ris}.
	This is the first work to comprehensively answer the channel shaping question and fully unveil the potential of \gls{bd}-\gls{ris} in multi-antenna systems.

	Fourth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} achievable rate maximization with a local-optimal approach via \gls{ao} and a low-complexity approach over channel shaping.
	The former updates active and passive beamforming by eigenmode transmission and \gls{rcg} algorithm, respectively.
	The latter suboptimally decouples both blocks, recasts the shaping problem as channel power maximization, and solves it iteratively in closed form.
	Interestingly, the gap in between vanishes as \gls{bd}-\gls{ris} evolves from diagonal to unitary.
	It suggests channel shaping helps to decouple the \gls{ris}-transceiver design and offers a promising low-complexity solution.
	This is the first work to discuss the implications of channel shaping on joint active-passive beamforming optimization.

	Fifth, extensive simulations reveal that the performance gain from \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions.
	In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\% and 270\% over diagonal \gls{ris} in $1 \times 1$ and $4 \times 4$ \gls{mimo} under independent Rayleigh fading, respectively.
	The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
	It emphasizes the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$ and $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ Hermitian and unitary matrices, respectively.
	% $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{arg}(\cdot)$ gives the argument of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis, we assume a lossless asymmetric network\footnote{While symmetric \gls{bd}-\gls{ris} (involving capacitors and inductors) is often considered in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2024,Zhou2023,Li2023d,Bartoli2023}, asymmetric \gls{bd}-\gls{ris} can be built over reconfigurable asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}.} without mutual coupling between scatter elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is a unitary matrix that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	That is, $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{H}$ and $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$.
	Note that diagonal (single-connected) and unitary (fully-connected) \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology have impacts on the scattering matrix.

>>>>

\begin{abstract}
	This paper investigates how a passive \gls{ris} can manipulate the \gls{mimo} point-to-point channel in terms of singular values and power gain.
	We depart from the widely-adopted diagonal phase shift model to a \gls{bd} architecture, which enables {cooperative} wave scattering thanks to in-group connections between elements.
	An efficient and universal \gls{bd}-\gls{ris} design framework is proposed over group-wise {geodesic} \gls{rcg} in the Stiefel manifold, which performs multiplicative rotational (instead of add-then-retract) updates for faster convergence.
	% For the singular value shaping problem, we numerically characterize their Pareto frontiers by weighted sum optimization via geodesic \gls{rcg}, which generalizes most relevant metrics (e.g., degree of freedom and condition number).
	% Many analytical singular value bounds are also derived for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	For the singular value problem, we numerically characterize their Pareto frontiers by geodesic \gls{rcg} and derive analytical bounds for rank-deficient channels and fully-connected (a.k.a. unitary) \gls{ris}.
	% For the singular value problem, we consider a numerical optimization approach that characterizes the Pareto frontiers as well as an analytical bounding approach in special (i.e., rank-deficient and fully-connected) scenarios.
	For the power gain problem, we provide a closed-form iterative solution by group-wise \gls{svd} that features even faster convergence.
	% Those channel shaping designs have a two-fold advantage.
	Those channel shaping designs have two major implications.
	First, they demystify the performance gain from off-diagonal entries in terms of {subchannel rearrangement} and {subspace alignment}.
	% help to demystify the gain from off-diagonal entries and reveal the shaping advantage of \gls{bd}-\gls{ris} in subchannel rearrangement and subspace alignment.
	Second, they inspire a two-stage suboptimal solution that decouples joint active and passive beamforming design.
	% \gls{ris}-transceiver optimization and suggest a low-complexity suboptimal solution based on channel power maximization.
	As a side product, we tackle the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by a local-optimal \gls{ao} approach and a two-stage shaping-inspired approach.
	Simulation results indicate that the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal towards unitary, while the relative gain of \gls{bd}-\gls{ris} increase with \gls{mimo} dimensions and decrease with spatial correlation.

@@

\begin{section}{Introduction}
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable metasurface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional multi-antenna techniques.
	The most popular \gls{ris} application is \emph{joint beamforming} design with transceivers, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	Since coherent scattering can be decomposed as equal-gain combining and transmission, \gls{ris} elements usually offer a squared asymptotic performance (e.g., second-order array gain and fourth-order power scaling law \cite{Zhao2022}) than transceiver antennas.
	% However, a large-dimensional \gls{ris} is generally required to compensate the power loss from double fading.
	On the other hand, \gls{ris} can also be employed for \emph{backscatter modulation} by periodically switching its reflection pattern.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) that can be either integrated into the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022}, or exploited by surrounding low-power devices for energy-efficient uplink communications \cite{Liang2020,Zhao2022a,Yang2024}.

	Despite the great potentials and fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape \gls{mimo} channels?}
	The answer depends heavily on the scattering model and hardware architecture.
	Most existing works assumed that each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	This ideally translates to a diagonal scattering matrix with unit-magnitude entries, which only controls the phase of the scattered waves but cannot redistribute their amplitude.
	The concept was soon generalized to \gls{bd}-\gls{ris} \cite{Shen2020a} which physically groups adjacent elements using passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices respectively.}
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate losslessly within the circuit and depart partially from any element in the same group.
	It can thus control both the amplitude and phase of the scattered wave without power loss, generalizing the scattering matrix to block-diagonal with unitary blocks.
	Such a powerful model can be realized at reduced hardware complexity using tree- and forest-connected architectures using graph theory \cite{Nerini2024}.
	\gls{bd}-\gls{ris} can also function in hybrid transmitting-and-reflecting mode \cite{Li2023b} and multi-sector mode \cite{Li2023c} for full-space coverage and multi-user support.
	Many practical design challenges have been addressed including channel estimation \cite{Li2024}, mutual coupling \cite{Li2023f}, and wideband modelling \cite{Li2024a}.
	Its superiority has been studied extensively in \gls{siso} and \gls{miso} systems in terms of single-user \gls{snr} maximization \cite{Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and multi-user \gls{wsr} maximization \cite{Fang2023,Zhou2023,Li2023c,Soleymani2024}.
	When it comes to \gls{mimo}, most existing works are limited to very specific scenarios.
	For example, the authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design under the assumption of unitary \gls{ris} without direct link.
	Received power maximization problem is extended to \gls{mimo} for continuous-valued \cite{Nerini2023} and discrete-valued \cite{Nerini2023b} scattering matrices, but they only considered single-stream transmission with given precoder and combiner (which is equivalent to \gls{siso}).
	Transmitter-side \gls{bd}-\gls{ris} design over statistical \gls{csi} has been introduced to massive \gls{mimo} networks for improved spectral efficiency \cite{Mishra2024}, but was again constrained to unitary \gls{ris} without direct link.
	A practical frequency-dependent \gls{bd}-\gls{ris} model has been recently proposed for multi-band multi-cell \gls{mimo} networks to facilitate practical deployments \cite{Sena2024}.
	Although the results are insightful, it remains unclear how the off-diagonal entries contribute to channel shaping and whether the relative gain of \gls{bd}-\gls{ris} over diagonal \gls{ris} scales with \gls{mimo} dimensions.
	% Therefore, \gls{bd}-\gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	Channel shaping is different from passive beamforming since the \gls{ris} is exploited as a stand-alone device to modify the inherent properties of the channel itself, instead of optimizing for a specific performance metric.
	This not only allows one to decouple the joint \gls{ris}-transceiver design, but also unveils the fundamental limits of wave manipulation and provides an all-in-one tool for various wireless applications.
	Relevant designs can be classified into two categories:
	\begin{itemize}
		\item \emph{Singular value centric:} Focus on the singular values of the equivalent channel matrix, which are sensitive to minor perturbations but precisely related to the capacity and diversity-multiplexing trade-off. It has been mainly studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}.
		\item \emph{Power centric:} Focus on the second-order characteristics of the equivalent channel matrix, which are useful in \gls{siso} and \gls{miso} but less informative in \gls{mimo}. It has been mainly studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and leakage interference \cite{Santamaria2023a}.
	\end{itemize}

	While the above works offer promising glimpses into the channel shaping potential of passive \gls{ris}, they neither suggest the achievable singular value region nor address the most general \gls{bd}-\gls{ris}-aided \gls{mimo} scenario with direct link and arbitrary group size.
	On the other hand, the great potential of \gls{bd}-\gls{ris} (especially in \gls{mimo}) has not been fully understood and an efficient design framework is still missing.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	First, we demystify the gain from off-diagonal entries of \gls{bd}-\gls{ris} in terms of subchannel rearrangement and subspace alignment.
	Subchannel rearrangement is a unique feature of cooperative scattering that allows each group to rearrange and recombine the associated forward and backward subchannels.
	This translates to a higher design flexibility and better shaping capability by exploiting the spatial diversity.
	On the other hand, phase matching in \gls{siso} generalizes to subspace alignment in \gls{mimo} and introduces a trade-off between the multiplicative backward-forward channel combination and the additive direct-indirect channel combination.
	As the \gls{mimo} dimensions expand, we show the benefits of \gls{bd}-\gls{ris} in subchannel rearrangement and the limitations of diagonal \gls{ris} in subspace alignment become increasingly apparent.
	This is the first paper to unveil and interpret the potential of \gls{bd}-\gls{ris} in multi-antenna systems.

	% where the \gls{bd}-\gls{ris} demonstrates a superior control on the multiplicative backward-forward channels and additive direct-indirect channels.
	% Both potentials

	Second, we exploit the Riemannian geometry of the Stiefel manifold and propose an efficient \gls{bd}-\gls{ris} design framework based on geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	This method modified from \cite{Abrudan2008,Abrudan2009} not only features lower complexity and faster convergence than general non-geodesic approach \cite{Absil2009,Pan2022d}, but also works for arbitrary group size and any smooth optimization problem.
	Specifically, block-wise multiplicative rotational updates are performed along the geodesics of the Stiefel manifold and compactly evaluated as the exponential map \cite{Edelman1998}.
	By exploiting the inherent structure of unitary matrices, this method avoids retractions from the Euclidean space and improves the computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

	Third, we quantify the capability of a \gls{bd}-\gls{ris} to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The {Pareto frontiers} are characterized by optimizing the {weighted sum of singular values}, where the weights can be positive, zero, or negative.
	This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The resulting achievable singular value region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and fully-connected (a.k.a. unitary) \gls{ris}.
	This is the first work to comprehensively answer the channel shaping capability question from a Pareto perspective.

	Fourth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} achievable rate maximization problem with two beamforming solutions: a local-optimal approach via \gls{ao} and a low-complexity approach over channel shaping.
	The former iteratively updates active beamforming by eigenmode transmission and passive beamforming by geodesic \gls{rcg} until convergence.
	The latter suboptimally decouples the joint design into a channel power gain maximization subproblem and a conventional \gls{mimo} transmission subproblem, then propose a two-stage solution in closed form.
	Interestingly, the rate deficit from the latter diminishes as the \gls{ris} evolves from diagonal to unitary.
	It suggests that channel shaping offers a promising path towards simplified and practical \gls{bd}-\gls{ris} (and transceiver) designs.
	% via \gls{bd}-\gls{ris} offers a promising path towards simplified and practical design.
	% offers a promising path towards simplified and practical desi
	% It suggests channel shaping helps to decouple the joint \gls{ris}-transceiver design and offers a promising low-complexity solution.
	% This is the first work to discuss the implications of channel shaping on joint active-passive beamforming optimization.

	% Fifth, extensive simulations reveal that the performance gain from \gls{bd}-\gls{ris} increases with group size and \gls{mimo} dimensions.
	% In terms of channel power, fully-connected \gls{bd}-\gls{ris} boosts up to 62\% and 270\% over diagonal \gls{ris} in $1 \times 1$ and $4 \times 4$ \gls{mimo} under independent Rayleigh fading, respectively.
	% The superiority stems from stronger \emph{subchannel rearrangement} and \emph{subspace alignment} capabilities empowered by in-group cooperation.
	% It emphasizes the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems.

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{H}^{n \times n}$ and $\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ Hermitian and unitary matrices, respectively.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{arg}(\cdot)$ gives the argument of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^\dagger$ $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), Moore-Penrose inverse, $r$-th iterated point, and stationary point, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$, $\lVert \cdot \rVert$, and $\lVert \cdot \rVert _\mathrm{F}$ denote the absolute value, Euclidean norm, and Frobenius norm, respectively.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis and explore the performance limits, we assume a lossless asymmetric network without mutual coupling between scattering elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is block-unitary
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is the $g$-th unitary block (i.e., $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$) that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	Note that diagonal and unitary \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the radiation pattern and circuit topology should be modelled in the scattering matrix.
